---
title: "Instroduction to R and Rstudio"
subtitle: "Session 5: Visualisation"
output: html_document
---


# Session 5 Introduction

The plots we produce will aim to present a large volume of information in graphical format. The ideal application would be for a presentation
There are four plots we will produce which can be broken down into different variations:
- A scatterplot: (i) basic (ii) with regression line, (iii) with groups (iv) groups and splines
- A Diverging barchart (i) Binary data split, (ii) Likert data
- A Modified boxplot for a t-test (i) Boxplot, (ii) T-distribution
- A Cell means Plot for ANOVA (can be used for RM/ Independent measures), this can also be used for time series data
- A heatmap and stacked bar chart for Factor Analysis and Factor loadings, conceptually, aheatmap may be applied to any clustering procedure

##0 Options and packages
We will primarily use `ggplot2` from the `tidyverse`. The useful thing is we can just load in tidyverse and it will provide all the packages we need, although we will need a few more, cowplot and ggnet
```{r}
library(tidyverse)
```

### GGplot2 methods

First lets make some practice data
```{r}
practice  <- read.csv("exs5.csv")

practice # view the data frame
```

Now we need to change the Time variable to a factor
```{r}
practice$Time  <- factor(practice$Time)

practice #view the data frame
```

GGplot2 works by stating what features in a data frame you wish to visualise, then adding the appropriate graph layers using the `+` operator. 

#### `aes()`
Graphs can be saved as objects at any stage of a build
GGplot2 takes in a set of aesthetics using the `aes()` function. `aes()` can be called inside the `ggplot()` function.
If we wanted to graph the distribution of the calories variable:
```{r}
ggplot(data = practice,
       aes(x=Calories)) +
  geom_histogram(bins = 15)
```
If we wanted to visualise the proportion of counts in each level of the curriculum category:
```{r}
ggplot(data = practice, aes(x = Curriculum))+
  geom_bar()
```

If we wanted to do a box plot of differences by Curriculum:
```{r}
ggplot(data = practice, aes(x=Curriculum, y = Calories))+
  geom_boxplot()
```

#### Rm ANOVA
So lets say we want to plot  for a Repeated Measures ANOVA. If we perform a summary using tidyverse functions we can get useable summary data which we cna then plot
```{r}
practice %>%
  group_by(Curriculum, Time) %>%
  summarise(Mean = mean(Calories), 
            StDev = sd(Calories), 
            n = n(), 
            StdError = StDev/sqrt(n)) ->> SummaryPractice
SummaryPractice
```

We can now use the `SummaryPractice` data frame for a cell-means interaction plot
```{r}
ggplot(data = SummaryPractice, aes(y= Mean, x = Time, group = Curriculum, colour=Curriculum))+
  geom_errorbar(aes(ymin=Mean - StdError, 
                    ymax=Mean + StdError), 
                width=.8, 
    position=position_dodge(0.05))+
  geom_line(size = 1) +
   geom_point(size = 5,
              fill='white',
             shape = 21)
```
We can also change the theme of the plot:
```{r}
ggplot(data = SummaryPractice, aes(y= Mean, x = Time, group = Curriculum, colour=Curriculum))+
  geom_errorbar(aes(ymin=Mean - StdError, 
                    ymax=Mean + StdError), 
                width=.8, 
    position=position_dodge(0.05))+
  geom_line(size = 1) +
   geom_point(size = 5,
              fill='white',
             shape = 21)+
  theme_classic()
```

We can also save the plot as an object
```{r}
ggplot(data = SummaryPractice, aes(y= Mean, x = Time, group = Curriculum, colour=Curriculum))+
  geom_errorbar(aes(ymin=Mean - StdError, 
                    ymax=Mean + StdError), 
                width=.8, 
    position=position_dodge(0.05))+
  geom_line(size = 1) +
   geom_point(size = 5,
              fill='white',
             shape = 21)+
  theme_classic() ->> Cell_MeansPlot1
```

```{r}
# Save as a .pdf file
#pdf("Cell_Means1.pdf")
Cell_MeansPlot1
#dev.off()
```
We can also view changes in time according to person by going back to the original data frame and using the `facet_grid()` layer:
```{r}
ggplot(data = practice, aes(y= Calories, x = Time , group = Person, colour=Person))+
  geom_line(size = 1) +
   geom_point(size = 5,
              fill='white',
             shape = 21)+
  facet_grid(~Person)+
  theme_classic()
```

```{r}
 aov(Calories ~ (Curriculum*Time) + Error(Person/(Time)), data = practice)
```

#ADVANCED GRAPHING

##1 Visualising a t-test
Taking the t-test data from session 2 example markdown file
```{r}
# Create long format data frame
df.long   <- data.frame(Response  = c(01,
                                      02, 02, 
                                      03, 03, 03, 
                                      04, 04, 04, 04,
                                      05, 05, 05, 05, 05, 
                                      06, 06, 06, 06, 06, 06,
                                      07, 07, 07, 07, 07, 07, 07, 
                                      08, 08, 08, 08, 08, 08,
                                      09, 09, 09, 09, 09,
                                      10, 10, 10, 10, 
                                      11, 11, 11, 
                                      12, 12,
                                      13,
                                      04, 
                                      05, 05,
                                      06, 06, 06, 
                                      07, 07, 07, 07, 
                                      08, 08, 08, 08, 08,
                                      09, 09, 09, 09, 09, 09, 
                                      10, 10, 10, 10, 10, 10, 10,
                                      11, 11, 11, 11, 11, 11, 
                                      12, 12, 12, 12, 12,
                                      13, 13, 13, 13, 
                                      14, 14, 14, 
                                      15, 15, 
                                      16),
                        Condition = c("M",
                                      "M", "M",
                                      "M", "M", "M",
                                      "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M",
                                      "M", "M", "M",
                                      "M", "M",
                                      "M",
                                      "F",
                                      "F", "F",
                                      "F", "F", "F",
                                      "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F",
                                      "F", "F", "F",
                                      "F", "F",
                                      "F"))
df.long
```
So now we build ur ttest:
```{r}
t.test(Response ~ Condition, data = df.long, var.equal=TRUE)
```

Now that we have the data we can begin to build the plot:
```{r}
qt(.95, 96)
pt(1.661, 96)
```

First plot the t distribution with a df of 96
Rejection Regions for the t-distribution with a df of 96:
```{r}
Rejection_Regions  <- data.frame(Alpha=c("5%", "5%", "1%", 
                                         "1%", "0.1%", "0.1%"),
                                 Rejection=c(qt(.95, df=96),
                                               qt(0.05, df = 96),
                                               qt(0.99, df = 96),
                                               qt(0.01, df = 96),
                                               qt(0.999, df = 96),
                                               qt(0.001, df = 96)))
Rejection_Regions  #1.660
```
Use the `dt()` function for density of t distribution, function.
```{r}
stats::dt(x = c(-5, 5), df= 96)
```

Plot the t-distribution with 96 df

```{r}
df   <- 96
p    <- 0.001
gg   <- data.frame(x=seq(-4,4,0.1))
gg$y <- dt(gg$x,df)
```

```{r}
tdist  <-  data.frame(x = c(-4, 4))
```

Visualise the t-test rejection region:
```{r}
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
        stat_function(fun = dt, args = list(df = 96))+
  geom_hline(yintercept = 0.0)+
  geom_area(stat = "function",args=list(df =96), fun = dt, fill = "tomato3", xlim = c(1.66, 4))+
  geom_area(stat = "function",args=list(df =96), 
            fun = dt, fill = "tomato3", xlim = c(-1.66, -4))+
  ggtitle(expression(paste("T-distribution ", alpha, " = 0.05,", italic(" df "), " = 96")))+
  scale_x_continuous(breaks = c(-4, -3,-1.66, -1, 0, 1, 1.66, 3, 4))+
  annotate('text', x = 3, y = .2, label= "5% Upper Rejection Region" )+
  annotate('text', x = -3, y = .2, label= "95% Lower Rejection Region" )+
  xlab("t")+
  theme_classic()->> reject2
reject2
```

```{r}
#pdf("rejection2.pdf")
reject2
#dev.off()
```

Add the coord_flip() layer to rotate the plot
```{r}
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
        stat_function(fun = dt, args = list(df = 96))+
  geom_hline(yintercept = 0.0)+
  geom_vline(xintercept = 0, linetype='dashed', colour='tomato3', size=2)+
  geom_vline(xintercept = -3.177308, colour='dodgerblue4', linetype='dashed', size=2)+
  geom_vline(xintercept = -1.660881, colour='tomato3')+
  geom_area(stat = "function",args=list(df =96), fun = dt, fill = "tomato3", xlim = c(1.66, 4))+
  geom_area(stat = "function",args=list(df =96), fun = dt, fill = "tomato3", xlim = c(-1.66, -4))+
  scale_x_continuous(limits = c(-10.45, 7))+
  theme_void() +
  theme(panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA))+
  coord_flip() ->> t.distplt

t.distplt
```


```{r}
ggplot(df.long, aes(Condition, Response, colour=Condition, fill=Condition))+
  geom_violin(fill='white',
              colour='grey22',
              size = 2)+
  geom_dotplot(stackdir = 'center', 
               binaxis = "y", 
               colour="grey22",
               dotsize = 1, 
               binwidth = .5)+
  geom_hline(yintercept=mean(df.long$Response[df.long$Condition=="F"]),
             colour='tomato3',
             linetype='dashed',
             size = 2)+
  geom_hline(yintercept=mean(df.long$Response[df.long$Condition=="M"]),
             colour = 'dodgerblue4',
             linetype='dashed', 
             size=2)+
  geom_point(data = df.summary, aes(x=Condition, y=mean), 
             shape=23, 
             size=8,
             fill="grey22",
             alpha=.8,
             stroke=1.5,
             show.legend = FALSE)+
  scale_fill_manual(values=c("tomato", "dodgerblue2"))+
  theme(legend.position=c(0.1, 0.1),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) ->> plt.data2

plt.data2

```

```{r}
# Plot panel
library(cowplot)
```

```{r}
# correct plot
#pdf("dotpltT3.pdf")
plot_grid(plt.data2, t.distplt, rel_widths = c(4,1))# Almost there
#dev.off()
```

Now we can plot the distribution density of the `Female` condition, with a lower rejection region (0.05 % of the distribution)
```{r}
# Create Normalised values for normal curve
xfit <- seq(min(df.long$Response[df.long$Condition=="F"]),
            max(df.long$Response[df.long$Condition=="F"]),
            length=40)

yfit <- dnorm(xfit,
              mean = mean(df.long$Response[df.long$Condition=="F"]),
              sd = sd(df.long$Response[df.long$Condition=="F"]))
```


```{r}
ggplot()+
  geom_line(aes(x=xfit, y = yfit),                   # Input x and y data
            colour="tomato",                         # line colour
            size = 1.5)+
  geom_vline(xintercept = mean(df.long$Response[df.long$Condition=="F"]),    # x axis intercept
             size = 1.5,
             colour= "tomato",                       # Colour
             linetype = "dashed")+
  scale_x_continuous(limits = c(0,16))+
coord_flip()+                                      # Rotate the plot
  theme_void() ->> dist.data

dist.data
```

Now using `cowplot` we can combine the two plots with a horizontal alignment, note that this works becasue we have set the x axis limits for the distribution plot to match the limits on the dotplot


Alternatively:
Extract a summary table and save as an object:
```{r}
df.long %>%
  group_by(Condition) %>%
  summarise(mean=mean(Response), st.dev=sd(Response), n=n()) %>%
  as_data_frame() ->> df.summary
df.summary  
```


####Reversing the plot for t-tests
The t-distribution portion of the plot can be visualised in reverse
##2 Visualising ANOVA
Here we will use the General Linear Model ANOVA, where the `aov()` function is replaced with `lm()`
ANOVA data from session 3 practical, where we will also include a second factor Variable.

```{r}
ggplot(df.long, aes(Condition, Response, colour=Condition, fill=Condition))+
  geom_boxplot(fill='white',
               colour='grey22',
               size = 2,
               notch = TRUE)+
  geom_dotplot(stackdir = 'center', 
               binaxis = "y", 
               colour="grey22",
               dotsize = 1, 
               binwidth = .5)+
  geom_hline(yintercept=mean(df.long$Response[df.long$Condition=="F"]),
             colour='tomato3',
             linetype='dashed',
             size = 2)+
  geom_hline(yintercept=mean(df.long$Response[df.long$Condition=="M"]),
             colour = 'dodgerblue4',
             linetype='dashed', 
             size=2)+
  geom_point(data = df.summary, aes(x=Condition, y=mean), 
             shape=23, 
             size=8,
             fill="grey22",
             alpha=.8,
             stroke=1.5,
             show.legend = FALSE)+
  scale_fill_manual(values=c("tomato", "dodgerblue2"))+
  theme(legend.position=c(0.1, 0.1),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) ->> plt.data3

plt.data3
```

Now we can reverse the scale length to visualise an upper rejection region
```{r}
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
        stat_function(fun = dt, args = list(df = 96))+
  geom_hline(yintercept = 0.0)+
  geom_vline(xintercept = 0, linetype='dashed', colour='dodgerblue4', size=2)+
  geom_vline(xintercept = 3.177308, colour='tomato3', linetype='dashed', size=2)+
  geom_vline(xintercept = 1.660881, colour='dodgerblue4')+
  geom_area(stat = "function",args=list(df =96), 
            fun = dt, fill = "dodgerblue4", xlim = c(1.66, 4))+
  geom_area(stat = "function",args=list(df =96), 
            fun = dt, fill = "dodgerblue4", xlim = c(-1.66, -4))+
  scale_x_continuous(limits = c(-7.5, 10.45))+
  theme_void() +
  theme(panel.background = element_rect(fill = "transparent", colour = NA), 
        plot.background = element_rect(fill = "transparent", colour = NA))+
  coord_flip() ->> t.distplt2

t.distplt2
```

```{r}
# correct plot
#pdf("dotpltT5.pdf")
plot_grid(plt.data3, t.distplt2, rel_widths = c(4,1))# Almost there
#dev.off()
```

```{r}
ggplot(df.long, aes(Condition, Response, colour=Condition, fill=Condition))+
  geom_dotplot(stackdir = 'center', 
               binaxis = "y", 
               colour="grey22", 
               binwidth = 1,
               dotsize = .5)+
  geom_hline(yintercept=mean(df.long$Response[df.long$Condition=="F"]),
             colour='tomato3',
             linetype='dashed',
             size = 2)+
  geom_hline(yintercept=mean(df.long$Response[df.long$Condition=="M"]),
             colour = 'dodgerblue4',
             linetype='dashed', 
             size=2)+
  geom_point(data = df.summary, aes(x=Condition, y=mean), 
             shape=23, 
             size=8,
             fill="grey22",
             alpha=.8,
             stroke=1.5,
             show.legend = FALSE)+
  scale_fill_manual(values=c("tomato", "dodgerblue2"))+
  theme_classic()+
  theme(legend.position=c(0.1, 0.1),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) ->> plt.data4

plt.data4

```

```{r}
# correct plot
#pdf("dotpltT6.pdf")
plot_grid(plt.data4, t.distplt, rel_widths = c(4,1))# Almost there
#dev.off()

```


##3 Visualising Psychometric Data
Psychometric Data from Session 4
We can also plot the data with a correlogram/heatplot with stacked bar chart to indicate loading qualitatively
```{r}
load <- read.csv("LoadingsB5.csv") # Load the csv file
load # O9 highest loading is 0.39 I have suppressed <0.4
```

Recode the items to make plotting more uniform later on
```{r}
library(tidyverse)
load$Item <- factor(load$Item)
#str(load$Item)
#levels(load$Item)

load$Item <- recode_factor(load$Item, 
                           O1 = "O01", O2 = "O02", O3 = "O03", O4 = "O04", O5 = "O05", 
                           O6 = "O06", O7 = "O07", O8 = "O08", O9 = "O09", O10 = "O10", 
                           C1 = "C01", C2 = "C02", C3 = "C03", C4 = "C04", C5 = "C05", 
                           C6 = "C06", C7 = "C07", C8 = "C08", C9 = "C09", C10 = "C10",
                           E1 = "E01", E2 = "E02", E3 = "E03", E4 = "E04", E5 = "E05", 
                           E6 = "E06", E7 = "E07", E8 = "E08", E9 = "E09", E10 = "E10",
                           A1 = "A01", A2 = "A02", A3 = "A03", A4 = "A04", A5 = "A05", 
                           A6 = "A06", A7 = "A07", A8 = "A08", A9 = "A09", A10 = "A10",
                           N1 = "N01", N2 = "N02", N3 = "N03", N4 = "N04", N5 = "N05", 
                           N6 = "N06", N7 = "N07", N8 = "N08", N9 = "N09", A10 = "A10") 
levels(load$Item)
```

```{r}
# Change names for factors
colnames(load)  <-  c("Item", "Extraversion", "Neuroticism", 
                      "Agreeableness", "Conscientiousness",
                      "Openness")
load
```


```{r}
# Now reshape the data set for ggplot2
library(reshape2)

loadings.m <- melt(load, id="Item", 
                   measure=c("Extraversion", "Neuroticism", 
                             "Agreeableness", "Openness",
                             "Conscientiousness"), 
                   variable.name="Factor", value.name="Loading")
loadings.m
```

This is a bar chart which shows the intensity of each factors loadings
Load the viridis library
```{r}
library(viridis)
```
#### Factor Loading Plot

```{r}
#For each test, plot the loading as length and fill color of a bar
# note that the length will be the absolute value of the loading but the 
# fill color will be the signed value, more on this below
ggplot(loadings.m, aes(Item, abs(Loading), fill=Loading)) + 
  facet_wrap(~ Factor, nrow=1) + #place the factors in separate facets
  geom_bar(stat="identity", colour='black') + #make the bars
  geom_hline(yintercept = 0.4, colour='blue')+
  coord_flip() + 
  ylab("Loading Strength") + #improve y-axis label
  scale_fill_viridis()+
  theme(panel.background = NULL, panel.border = element_blank(),
        strip.background =element_rect(fill="white"),
        strip.text.x = element_text(size = 8),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 6))+
  guides(fill=FALSE)->> LoadPlot


LoadPlot
```

With magma Colouring
```{r}
LoadPlot+  scale_fill_viridis(option="A")#magma
```


## Correlation matrix
```{r}
Ord <- c("O01","O02", "O03", "O04", "O05", "O06", "O07", "O08", "O09", "O10", 
         "C01", "C02", "C03", "C04", "C05", "C06", "C07", "C08", "C09", "C10",
         "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09", "E10", 
         "A01", "A02", "A03", "A04", "A05", "A06", "A07", "A08", "A09", "A10",
         "N01", "N02", "N03", "N04", "N05", "N06", "N07", "N08", "N09", "N10")
colnames(data_FA) <- Ord

efa_cor <- round(cor(data_FA),2)
corrs.m <- melt(efa_cor, id="Test", variable.name="Test2", value.name="Correlation")
corrs.m
#write.csv(corrs.m, "FullCormat.csv")
#corrs.m$Test2 <- reorder(corrs.m, rep(Ord, each=50)) 
summary(corrs.m)
```

```{r}
corrs.m2  <- read.csv("FullCormat.csv")
corrs.m2$x  <- NULL

melted_cormat2  <- read.csv("Cormatplot.csv")
melted_cormat2$X  <- NULL
```

Next create a correlation matrix of data

```{r}
library(grid)
p1 <- ggplot(data = melted_cormat2, aes(Var2, Var1, fill = value))+
 geom_tile(color = "black")+
  scale_fill_viridis( limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 8, hjust = 1),
       axis.text.y = element_text(size = 8),
       plot.margin = unit(c(3, 1, 0, 0), "mm"),
       panel.background = NULL)+
  coord_fixed()+
  guides(fill = FALSE) #omit unnecessary gradient legend
p1
# Note I removed the correlation values as needed
# reinsert the following line to produce correlation values
#   geom_text(aes(label = round(value, 2)), size=2.)
```

```{r}
p2_alt<- ggplot(data = corrs.m2, aes(Var2, Var1, fill = Correlation))+
 geom_tile(color = "black")+
  scale_fill_viridis( limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 1, size = 8, hjust = 1),
        axis.title.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size = 8),
       plot.margin = unit(c(3, 1, 0, 0), "mm"),
       panel.background = NULL)+
  coord_fixed()+
  guides(fill = FALSE)

p2_alt
```

```{r}
ggplot(data = corrs.m2, aes(Var2, Var1, fill = Correlation))+
 geom_tile(color = "black")+
  scale_fill_viridis( limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 1, size = 8, hjust = 1),
        axis.title.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size = 8),
       plot.margin = unit(c(3, 1, 0, 0), "mm"),
       panel.background = NULL)+
  coord_fixed()+
  scale_fill_viridis(option="D")#magma
```

```{r}

p2_alt + scale_fill_viridis(option="A")#magma

```

```{r}
ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "black")+
  theme_minimal()+ 
  scale_colour_gradient2(high = 'yellow', 
                         mid = "white",
                         low = 'purple', 
                         midpoint = 0, space = "Lab")+
 theme(axis.text.x = element_text(angle = 90, vjust = 1, 
    size = 8, hjust = 1),
    axis.text.y = element_text(size = 8))+
 coord_fixed()

```



```{r}
# Stacked Bar Chart for Factor loadings
p2 <- ggplot(loadings.m, aes(Item, abs(Loading), fill=Factor)) + 
  geom_bar(stat="identity", colour='black') + coord_flip() + 
  ylab("Loading Strength") + theme_bw(base_size=10) + 
  scale_fill_viridis(discrete = TRUE)+
  theme_classic()+
  #remove labels and tweak margins for combining with the correlation matrix plot
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 90),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank()) -> pload1


pload1
```

```{r}
pload2  <- pload1 + theme(legend.position = "none")
pload2
```

Change to Magma Colouring
```{r}
loadMagma  <- pload2 +  scale_fill_viridis(discrete=TRUE, option="A")#magma
```


```{r}
library(cowplot)
p2_Leg <- get_legend(pload1)
p2_Leg
```


Now panel `p1` (correlation plot) and `p2` (stacked bar-chart)
```{r}
library(gridExtra)
grid.arrange(p1_alt, pload2, ncol=2, widths=c(2, 1))

```

```{r}
# cowplot
plot_grid(p2_alt, pload2, ncol=2, rel_widths = c(2, 1), align="h") ->> ldplt2

```
```{r}
ldplt2
```

```{r}
#pdf("LoadingPlotHeatmap2.pdf")
ldplt2
#dev.off()
```